How Consistent are LLMs in Applying Handcrafted Language Features

### 2024-06-09 article/main.tex
update prompts. update plots
incorporate first results in article text

\usepackage{algorithmic}
\usepackage{algorithm} 
\usepackage{array}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{cite}
\usepackage{comment} 
\usepackage{enumitem}
\usepackage{epsfig} 
\usepackage{fancyhdr}
\usepackage[T1]{fontenc}
\usepackage{geometry} 
\usepackage{graphicx}
\usepackage[colorlinks]{hyperref} 
\usepackage[latin1]{inputenc}
\usepackage{interval}
\usepackage{multicol}
\usepackage{multirow} 
\usepackage{rotating}
\usepackage{setspace}
\usepackage{subfigure}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{url}
\usepackage{verbatim}
\usepackage{xcolor}
\usepackage{xstring}

@@

In order to perform our experiment we must first choose proper candidates.
Since the interaction with the model happens only via prompts, we may use models
with any type of licensing.
For our experiment we choose the top model from each of the highest scoring five
different major vendors listed on the Chatbot Arena~\cite{chiang2024chatbot}
leaderboard.
The final choice is documented in Table~\ref{table-llm}.

\begin{table}[ht]
    \setlength\tabcolsep{6pt}
    \centering
    \begin{tabular}{@{}lllll@{}}\toprule
        Name                      & Parameters  & Context Size & Elo Score & License Type        \\ \toprule
        OpenAI GPT-4o             & undisclosed & 128000       & 1287      & Proprietary         \\
        Google Gemini 1.5 Pro     & undisclosed & 1048576      & 1268      & Proprietary         \\
        Anthropic Claude 3 Opus   & undisclosed & 200000       & 1248      & Proprietary         \\
        Meta Llama 3-70B-Instruct & 70 Billion  & 8000         & 1208      & Llama 3 Community   \\
        Cohere Command R+         & 104 Billion & 128000       & 1189      & CC-BY-NC 4.0        \\
                                  &             &              &           & with Acceptable Use \\
                                  &             &              &           & Addendum            \\ \bottomrule
    \end{tabular}
    \caption{Experiment LLMs}\label{table-llm}
\end{table}

@@

Another HLF characteristic we have in mind is their level of abstraction.
Features such as the word count or number of punctuation mark are very concrete.
On the other side of the spectrum, features such as the Flesch-Kincaid grade
level~\cite{kincaid1975derivation} are highly abstract.
We divide HLFs in five categories ranging from very concrete to highly abstract
and choose 2 features from each category for a grand total of 10 HLFs.
Our final selection is available in Table~\ref{table-hlf}.

\begin{table}[ht]
    \setlength\tabcolsep{6pt}
    \centering
    \begin{tabular}{@{}lllll@{}}\toprule
        Name                              & Family           & Domain           & Abstraction     & Style Level\footnotemark \\ \toprule
        total unique words                & wordsent         & surface          & very concrete   & lexical                  \\
        total sentences                   & wordsent         & surface          & very concrete   & syntactic                \\ \bottomrule
        total number of named entities    & entity           & discourse        & concrete        & semantic                 \\
        total number of unique adjectives & partofspeech     & syntax           & concrete        & syntactic                \\ \bottomrule
        simple type token ratio           & typetokenratio   & lexico-semantics & regular         & pragmatic                \\
        number of verbs per word          & partofspeech     & surface          & regular         & syntactic                \\ \bottomrule
        corrected adjective variation     & lexicalvariation & lexico-semantics & abstract        & lexical                  \\
        corrected verb variation          & lexicalvariation & lexico-semantics & abstract        & lexical                  \\ \bottomrule
        Flesch-Kincaid grade level        & readformula      & surface          & highly abstract & pragmatic                \\
        Kuperman age of aquisition        & worddiff         & lexico-semantics & highly abstract & lexical                  \\ \bottomrule
    \end{tabular}
    \caption{Handcrafted Linguistic Features}\label{table-hlf}
\end{table}
\footnotetext{as described in the figure in~\cite{lugea2023stylistics}, chapter 1, page 7}

@@

\begin{table}[ht]
    \setlength\tabcolsep{6pt}
    \centering
    \begin{tabular}{@{}lllll@{}}\toprule
        Name              & Task                       & Row Count & Content Type & Reference                     \\ \toprule
        Yelp Reviews Full & Sentiment Classification   & 700000    & Reviews      & \cite{yelp2015neurips}        \\
                          & Text Style Transfer        & 700000    &              &                               \\
                          & Text Classification        & 700000    &              &                               \\ \midrule
        CNN / DailyMail   & Summarization              & 300000    & News         & \cite{cnndm2015}              \\
                          & Question Answering         &           & Stories      & \cite{cnndm2017}              \\
                          & Text Generation            &           &              &                               \\ \midrule
        HANNA             & Automatic Story Generation & 1056      & Stories      & \cite{hanna2024}              \\ \midrule
        StylePTB          & Text Style Transfer        & 60000     & Sentences    & \cite{lyu-etal-2021-styleptb} \\ \bottomrule
    \end{tabular}
    \caption{Experiment LLMs}\label{table-ds}
\end{table}

@@

The Human-ANnotated NArratives (HANNA) dataset is interesting because it
contains stories generated by LLMs. Since LLM generated stories can be assessed
just as easily as human generated stories by using HLFs, we thought it would be
a welcome addition to run our experiment for non-human writers, too.

Finally, StylePTB is a dataset containing fine grained stylistic changes at the
sentence level. We use this dataset to see how well LLMs do in scenarios
involving HLFs at the sentence level (for example, part of speech HLFs).

\subsection{Input Prompt and Text}\label{input-text}

The input prompts are derived from the templates available in the Git repository
that accompanies this article~\cite{olar2024experimentcode}.
We provide the LLM with a system prompt that contains instructions concerning
all HLFs.

@@

\subsection{Graph Interpretation}

Handcrafted linguistic features are integer or floating point values that we
can compute on text.
All our graphs show the HLF on the vertical axis, while the horizontal axis is
reserved for the trial number.
The plots lend themselves to easy interpretation based on some simple rules.
We plot the HLF values for the text generated from baseline prompts as a
continuous grey line.
The similar values obtained by using HLF enhanced prompts are plotted using a
continuous grenat line.
To highlight the `tolerance' obtained by adding and subtracting the variance of
the HLF on the benchmark dataset, we plot light gray areas around the baseline
and HLF-enhanced plot lines.
The minimum and maximum HLF on the benchmark dataset are plotted using dotted
black lines.
A continuous black line designates the average HLF on the benchmark dataset.

Following this description of how to interpret the graphs, we present sample
findings using the features for which we've observed the best and the worst
compliance by the LLM.
These findings were all made on the <DECIDE LATER> dataset.

@@

Yelp Results on the best and worst performing metric.
The rest of the metrics in an Appendix.

>>>>

\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{array}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{booktabs}
\usepackage[justification=centering,font=scriptsize]{caption}
\usepackage{cite}
\usepackage{comment}
\usepackage{enumitem}
\usepackage{epsfig}
\usepackage{fancyhdr}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage[colorlinks]{hyperref}
\usepackage{inconsolata}
\usepackage[latin1]{inputenc}
\usepackage{interval}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{setspace}
\usepackage{subfigure}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{url}
\usepackage{verbatim}
\usepackage{xcolor}
\usepackage{xstring}

@@

For our experiment we will use some of the best large language models available.
We trust the competence judgements issued in the the Chatbot Arena~\cite{chiang2024chatbot}
for each LLM.
In the interest of covering more ground, we choose large language models from
different vendors.
Our choice includes four top performing LLMs and one model that anyone would be
able to run on their own hardware.
Our final choice is documented in Table~\ref{table-llm}.

\begin{table}[ht]
    \setlength\tabcolsep{6pt}
    \centering
    \begin{tabular}{@{}lllll@{}}
        \toprule
        Name                    & Parameters  & Context Size & Elo Score & License Type        \\ \toprule
        OpenAI GPT-4o           & undisclosed & 128000       & 1287      & Proprietary         \\
        Google Gemini 1.5 Pro   & undisclosed & 1048576      & 1268      & Proprietary         \\
        Anthropic Claude 3 Opus & undisclosed & 200000       & 1248      & Proprietary         \\
        Meta Llama 3-8B         & 8 Billion   & 8000         & 1208      & Llama 3 Community   \\
        Cohere Command R+       & 104 Billion & 128000       & 1189      & CC-BY-NC 4.0        \\
                                &             &              &           & with Acceptable Use \\
                                &             &              &           & Addendum            \\ \bottomrule
    \end{tabular}
    \caption{Experiment LLMs}\label{table-llm}
\end{table}

@@

We can also describe HLFs in terms of how abstract the instruction to conform to
them is.
For example, `write 100 words' is a more concrete instruction than `write as if
you were in junior highschool'~\cite{kincaid1975derivation}.
We divide HLFs in five categories ranging from very concrete to highly abstract
and choose 2 features from each category for a grand total of 10 HLFs.
Table~\ref{table-hlf} showcases our selection of features.

\begin{table}[ht]
    \setlength\tabcolsep{6pt}
    \centering
    \begin{tabular}{@{}lllll@{}}
        \toprule
        Name                              & Family           & Domain           & Abstraction     & Style Level\footnotemark \\ \toprule
        total words                       & wordsent         & surface          & very concrete   & lexical                  \\
        total sentences                   & wordsent         & surface          & very concrete   & syntactic                \\ \bottomrule
        total number of unique adjectives & partofspeech     & syntax           & concrete        & semantic                 \\
        total number of unique adjectives & partofspeech     & syntax           & concrete        & syntactic                \\ \bottomrule
        simple type token ratio           & typetokenratio   & lexico-semantics & regular         & pragmatic                \\
        number of verbs per word          & partofspeech     & surface          & regular         & syntactic                \\ \bottomrule
        corrected adjective variation     & lexicalvariation & lexico-semantics & abstract        & lexical                  \\
        corrected verb variation          & lexicalvariation & lexico-semantics & abstract        & lexical                  \\ \bottomrule
        Flesch-Kincaid grade level        & readformula      & surface          & highly abstract & pragmatic                \\
        Kuperman age of aquisition        & worddiff         & lexico-semantics & highly abstract & lexical                  \\ \bottomrule
    \end{tabular}
    \caption{Handcrafted Linguistic Features}\label{table-hlf}
\end{table}
\footnotetext{as described in the figure in~\cite{lugea2023stylistics}, chapter 1, page 7}

Each of these features is well documented in literature~\cite{lftk-2023}.

@@

\begin{table}[ht]
    \setlength\tabcolsep{6pt}
    \centering
    \begin{tabular}{@{}lllll@{}}
        \toprule
        Name              & Task                     & Row Count & Content Type & Reference              \\ \toprule
        Yelp Reviews Full & Sentiment Classification & 700000    & Reviews      & \cite{yelp2015neurips} \\
                          & Text Style Transfer      & 700000    &              &                        \\
                          & Text Classification      & 700000    &              &                        \\ \midrule
        CNN / DailyMail   & Summarization            & 300000    & News         & \cite{cnndm2015}       \\
                          & Question Answering       &           & Stories      & \cite{cnndm2017}       \\
                          & Text Generation          &           &              &                        \\ \midrule
    \end{tabular}
    \caption{Experiment LLMs}\label{table-ds}
\end{table}

@@

We use the Huggingface ecosystem to load these
datasets~\cite{lhoest-etal-2021-datasets}.
Because of the statistical nature of the experiment and in the interest of
expediency, we only use the first thousand texts from the test split of each
dataset.

\subsection{User Prompt and Text}\label{input-text}

This experiment consists of instructing various large language models to reword
an input text so that it conforms to a bunch of handcrafted linguistic features.
The input text is provided via a user prompt.
We provide two user prompts for each dataset.
These user prompts can be freely examined online in the
\texttt{\href{https://github.com/koosie0507/llm-tst-consistency/tree/main/data}{data}}
folder of the experiment's Git repository~\cite{olar2024experimentcode}.

For the CNN/DailyMail dataset, the user prompt from outside the dataset is an
extract from the classic text`The Life and Opinions of Tristram Shandy,
Gentleman', by Laurence Sterne~\cite{sterne2003life}.
We judge this text to be far apart in both style and exhibited linguistic
features from the CNN/DailyMail dataset contents.
For the Yelp dataset, our chosen user prompt is Oscar Wilde's `Sonnet to
Liberty'~\cite{wilde1909poems}.
This text is vastly different from anything in the Yelp review dataset.

The instructions for the LLM are compiled from the Jinja templates available in
the \texttt{\href{https://github.com/koosie0507/llm-tst-consistency/tree/main/data}{templates}}
folder of the experiment's Git repository~\cite{olar2024experimentcode}.
This folder includes both the main instruction prompts for the two scenarios
that we cover in this experiment (`\texttt{prompt\_1}' and `\texttt{prompt\_2}')
and the templates for generating the instructions for each HLF.

@@

\subsection{Interpretation Conventions}

Handcrafted linguistic features are integer or floating point values that we
can compute on text.
All our graphs show the HLF on the vertical axis, while the horizontal axis is
reserved for the trial number.
We plot the HLF values for the text generated from baseline prompts as a
continuous grey line.
The similar values obtained by using HLF enhanced prompts are plotted using a
continuous grenat line.
To highlight the `tolerance' obtained by adding and subtracting the variance of
the HLF on the benchmark dataset, we plot light gray areas around the baseline
and HLF-enhanced plot lines.
The minimum and maximum HLF on the benchmark dataset are plotted using dotted
black lines.
A continuous black line designates the average HLF on the benchmark dataset.

Our results tables highlight differences between the generated text's HLF values
obtained by using the baseline and the HLF prompts.
We are concerned with how different the plots of each HLF are from the
continuous black line representing the average HLF on the benchmark dataset and
how different they are from each other.

We use the two-sampled Kolmogorov-Smirnov test to determine whether the baseline
results are different from the HLF-enhanced results.
We consider that any $p-value \le 0.05$ means that the distributions are
different.

To determine which of the two is closer to the target, we compute the area
between each result and the target.
Another way to measure the geometric closeness to the target is by computing the
Euclidian norm of the vectorial difference between result and target.
Let us denote with $A_b$ and $A_h$ the areas between baseline and target and HLF
instruction result and target, respectively.
Let us similarly denote with $N_b$ and $N_h$ the Euclidian norms.
Finally, we define $Diff_A = A_b - A_h$ and $Diff_N = N_b - N_h$

The tables that sum up the experiment results show $Diff_A$ and $Diff_N$ for
each model and we provide a result table for each dataset and prompt variant in
the online repository of the experiment.
For each value we specify whether it is significant statistically according to
the Kolmogorov-Smirnov test.

In the next sections we only present the most relevant results of the
experiment.
To save space, we illustrate significant differences between the baseline and
the HLF instruction results in \textbf{bold} typeface.
Positive results are \underline{underlined} and negative results are
\textit{italicized}.

@@

The results on the CNN-DailyMail dataset showcased in
Table~\ref{table-cnn-dailymail-results} provide the most eloquent answer to this
question.
The results on the Yelp dataset are positive, albeit less convincing.
\begin{table}[!ht]
    \centering
    \resizebox{\textwidth}{!}{%
        \begin{tabular}{lllllllllll}
            model               & t\_word                      & t\_sent                    & n\_uverb                    & n\_uadj                    & simp\_ttr                 & a\_verb\_pw               & corr\_adj\_var            & corr\_verb\_var            & fkgl                    & a\_kup\_pw                \\
            \toprule
            claude3 $Diff_N$    & \underline{\textbf{623.16}}  & \underline{\textbf{24.74}} & \underline{\textbf{51.18}}  & \underline{\textbf{26.68}} & 0.06                      & -0.0                      & \underline{\textbf{0.46}} & \underline{\textbf{1.4}}   & \underline{1.44}        & \textit{\textbf{-1.66}}   \\
            claude3 $Diff_A$    & \underline{\textbf{1774.0}}  & \underline{\textbf{74.0}}  & \underline{\textbf{158.84}} & \underline{\textbf{81.2}}  & 0.2                       & -0.01                     & \underline{\textbf{1.88}} & \underline{\textbf{4.49}}  & \underline{5.01}        & \textit{\textbf{-4.09}}   \\
            gemini $Diff_N$     & \underline{\textbf{985.86}}  & \underline{\textbf{34.52}} & \underline{\textbf{100.36}} & \underline{\textbf{53.39}} & \underline{\textbf{0.52}} & 0.05                      & \underline{\textbf{2.62}} & \underline{\textbf{4.68}}  & \textit{-4.17}          & \textit{\textbf{-2.43}}   \\
            gemini $Diff_A$     & \underline{\textbf{2990.5}}  & \underline{\textbf{106.5}} & \underline{\textbf{320.34}} & \underline{\textbf{159.1}} & \underline{\textbf{1.5}}  & 0.09                      & \underline{\textbf{7.24}} & \underline{\textbf{13.15}} & \textit{-12.65}         & \textit{\textbf{-7.13}}   \\
            gpt $Diff_N$        & \underline{\textbf{125.87}}  & 1.46                       & 10.95                       & 10.77                      & 0.02                      & -0.01                     & 0.71                      & 0.81                       & \textit{\textbf{-1.72}} & 0.43                      \\
            gpt $Diff_A$        & \underline{\textbf{327.0}}   & 3.0                        & 29.0                        & 27.0                       & 0.08                      & -0.01                     & 1.82                      & 2.24                       & \textit{\textbf{-5.72}} & 0.44                      \\
            llama3\_8b $Diff_N$ & 171.31                       & 6.63                       & -9.8                        & 13.8                       & 0.04                      & -0.04                     & 1.03                      & -0.84                      & -0.11                   & -0.76                     \\
            llama3\_8b $Diff_A$ & 524.0                        & 20.0                       & -32.5                       & 43.5                       & 0.1                       & -0.08                     & 3.64                      & -2.85                      & 1.57                    & -2.29                     \\
            command\_r $Diff_N$ & \underline{\textbf{1024.78}} & \underline{\textbf{39.67}} & \underline{\textbf{81.58}}  & \underline{\textbf{41.46}} & \underline{\textbf{0.46}} & \underline{\textbf{0.03}} & \underline{\textbf{2.67}} & \underline{\textbf{3.78}}  & \underline{1.4}         & \underline{\textbf{0.14}} \\
            command\_r $Diff_A$ & \underline{\textbf{3132.5}}  & \underline{\textbf{120.5}} & \underline{\textbf{246.84}} & \underline{\textbf{132.2}} & \underline{\textbf{1.36}} & \underline{\textbf{0.06}} & \underline{\textbf{7.88}} & \underline{\textbf{11.21}} & \underline{3.9}         & \underline{\textbf{1.03}} \\
            \bottomrule
        \end{tabular}%
    }
    \caption{CNN-DailyMail Results}
    \label{table-cnn-dailymail-results}
\end{table}

Command-R+ was the best performer by far, while Llama3 (8B) was predictibly the
worst at following HLF instructions.
Figure~\ref{fig-command-r-t-word} shows a positive result, where the result
obtained with the aid of HLF instructions is closer to the target.
Figure~\ref{fig-gpt-fkgl} shows a negative result, where the baseline is closer
to the target.
Figure~\ref{fig-llama-simp-ttr} shows an inconclusive result.

\begin{figure}[ht!]
    \centering
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_1/prompt_1-command_r-cnn_dailymail/prompt_1-command_r-cnn_dailymail_t_word.png}
        \caption{Command-R+\\Total Words}
        \label{fig-command-r-t-word}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_1/prompt_1-gpt-cnn_dailymail/prompt_1-gpt-cnn_dailymail_fkgl.png}
        \caption{GPT\\Flesch-Kincaid Grade Level}
        \label{fig-gpt-fkgl}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_1/prompt_1-llama3_8b-cnn_dailymail/prompt_1-llama3_8b-cnn_dailymail_simp_ttr.png}
        \caption[center]{Llama3-8B\\Simple TTR}
        \label{fig-llama-simp-ttr}
    \end{minipage}
\end{figure}
